% =============================================================================
% Reeling in Smart Attackers' Phishing Emails with Neural Networks
% IEEE conference (IEEEtran) two-column manuscript.
% Compile: pdflatex main && bibtex main && pdflatex main && pdflatex main
% Overleaf: upload this folder; ensure "IEEEtran" template is available (it is
% bundled in Overleaf's TeX Live).
% =============================================================================

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\scriptsize,
  breaklines=true,
  frame=single,
  framesep=2pt,
  framerule=0.3pt,
  xleftmargin=2pt,
  xrightmargin=2pt,
  aboveskip=4pt,
  belowskip=4pt,
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Reeling in Smart Attackers' Phishing Emails with Neural Networks\\
{\footnotesize \textsuperscript{}CSE 587 Final Project --- Group 5, Spring 2026}
}

\author{
\IEEEauthorblockN{Parker Davis}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{The Pennsylvania State University}\\
University Park, PA, USA \\
pqd5283@psu.edu}
\and
\IEEEauthorblockN{Dillon VanGilder}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{The Pennsylvania State University}\\
University Park, PA, USA \\
dsv5165@psu.edu}
\and
\IEEEauthorblockN{Emmanuel Adjei Domfeh}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{The Pennsylvania State University}\\
University Park, PA, USA \\
eza5399@psu.edu}
}

\maketitle

\begin{abstract}
Phishing email detection is typically benchmarked on datasets in which
malicious messages contain conspicuous surface cues (urgency keywords, raw
URLs, mismatched senders), and Transformer-based classifiers achieve near-99\%
accuracy on such test sets. A sophisticated (``smart'') attacker --- or,
increasingly, an LLM--generated phishing email --- can remove or paraphrase
those cues while preserving malicious intent, raising the question of whether
the high reported accuracy reflects robustness or surface-feature memorization.
We fine-tune \textsc{BERT}\cite{devlin2019bert} on the curated multi-source
phishing email corpus released by Champa~et~al.\cite{champa2024curated}
($N\!=\!131{,}171$) and evaluate on a held-out \emph{smart-attacker} test split
obtained by applying label-preserving edits to phishing messages: URL masking,
urgency-keyword removal, suspicious subject rewriting, synonym substitution,
and light character perturbation. We then retrain on training data augmented
with the same edits and re-evaluate. Each configuration is run with three
seeds and we report mean~$\pm$~std. Our findings are two-fold. First,
contrary to our initial hypothesis, vanilla \textsc{BERT-base} is essentially
\emph{perfectly} robust to our smart-attacker edits on this corpus: phishing
F1 changes by $-0.0001$ (0.9958~$\pm$~0.0002 $\rightarrow$
0.9957~$\pm$~0.0001) under the smart-attacker test --- within
seed-to-seed noise. We attribute this to redundancy in the corpus's phishing
signal and to BERT's distributed representation. Second, targeted augmentation
nevertheless produces a clean, statistically tight robustness gain
($+0.0020$ F1, 0.9957~$\pm$~0.0001 $\rightarrow$ 0.9977~$\pm$~0.0001) at
\emph{zero} cost on the standard test ($+0.0000$ F1).
We further provide a per-edit ablation showing that the augmented model
gains the most when phishing emails are subjected to character-level
perturbation (F1 = 0.9975~$\pm$~0.0001) --- evidence that augmentation
teaches the classifier to attend to deeper semantic features rather than
to specific token forms. We discuss how the same pipeline could be
re-evaluated against modern, LLM-generated phishing where larger baseline
drops are expected.
\end{abstract}

\begin{IEEEkeywords}
phishing detection, BERT, Transformers, robustness, data augmentation,
adversarial NLP
\end{IEEEkeywords}

% =============================================================================
\section{Introduction}
\label{sec:intro}

Phishing remains one of the most effective initial-access vectors in modern
cyberattacks. Despite years of user-awareness training, recent work shows that
sophisticated --- and especially \emph{LLM--generated} --- phishing emails
sustain alarming click rates~\cite{kim2025llms}. The same wave of generative
models that helps defenders also helps attackers: an LLM can rewrite a
suspicious message to sound natural, drop tell-tale urgency keywords, and
disguise a malicious link as an innocent-looking domain.

Mainstream phishing classifiers are typically Transformer-based: BERT and
RoBERTa fine-tunes routinely report 98--99\% accuracy on standard benchmarks
\cite{atawneh2023phishing,phishingbertroberta2025,indepthphishingdl2025}.
However, this performance is measured on data with strong, often artifactual,
phishing cues. When those cues are removed, classifier accuracy drops
substantially~\cite{adversarialphishing2023,explainabletransformer2025,robustllmphish2025}.

\paragraph*{Contribution.} This work investigates two questions:
\begin{enumerate}
  \item Do standard phishing detectors over-rely on superficial cues
        (URLs, urgency keywords, suspicious subject prefixes)?
  \item Can targeted, label-preserving data augmentation that simulates
        a ``smart attacker'' improve robustness without significantly
        degrading performance on the original test distribution?
\end{enumerate}

We instantiate the workflow in Fig.~\ref{fig:workflow}. Concretely, we
contribute (i) a label-preserving smart-attacker augmentation pipeline
(\S\ref{sec:augmentation}); (ii) a held-out smart-attacker evaluation
split; (iii) a per-edit ablation that localizes which surface cue the
classifier relies on; (iv) a multi-seed (n\,=\,3) experimental protocol
that distinguishes signal from seed noise; and (v) a fully reproducible
PyTorch~+~HuggingFace codebase that runs end-to-end in a single Colab
notebook.

\paragraph*{Summary of findings.} Our experiments do not support the
over-reliance hypothesis on the Champa~et~al.\ corpus: the baseline
classifier loses essentially zero F1 ($-0.0001 \pm 0.0001$) under the
smart-attacker test, well within seed noise. Augmentation nevertheless
produces a clean Pareto improvement, gaining $+0.0020$\,F1 on the
smart-attacker split at no measurable cost on the standard split, with
\texttt{char\_perturb} as the most impactful single edit in the per-edit
ablation. We argue these results reflect properties of the corpus and of
BERT's distributed representation, and we sketch how the same pipeline
could be re-applied to modern LLM-generated phishing where larger
baseline drops are expected.

\begin{figure}[t]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
  node distance=5mm and 5mm,
  font=\footnotesize,
  box/.style={draw, rounded corners=1pt, align=center,
              minimum width=24mm, minimum height=6mm, fill=blue!5},
  data/.style={draw, ellipse, align=center,
               minimum width=22mm, minimum height=6mm, fill=red!8},
  eval/.style={draw, rounded corners=1pt, align=center,
               minimum width=24mm, minimum height=6mm, fill=green!8},
  arr/.style={-Latex, thick}
]
\node[data] (raw) {Phishing dataset \\ \scriptsize{\cite{champa2024curated}}};
\node[box, below=of raw] (train) {Train BERT};
\node[eval, below=of train] (e1) {Eval (standard)};
\node[box, right=10mm of e1] (aug) {Augment data \\ \scriptsize{(smart attacker)}};
\node[eval, above=of aug] (e2) {Eval (smart attacker)};
\node[box, above=of e2] (rt) {Retrain on \\ augmented data};
\node[eval, right=10mm of rt] (e3) {Eval again \\ (both splits)};

\draw[arr] (raw)   -- (train);
\draw[arr] (train) -- (e1);
\draw[arr] (e1)    -- (aug);
\draw[arr] (aug)   -- (e2);
\draw[arr] (e2)    -- (rt);
\draw[arr] (rt)    -- (e3);
\end{tikzpicture}}
\caption{Three-stage experimental workflow. (1) Train BERT on the
Champa~et~al.\ corpus. (2) Evaluate on the standard test set and a
smart-attacker variant produced by applying label-preserving edits. (3)
Augment training data with the same edits, retrain, and re-evaluate.}
\label{fig:workflow}
\end{figure}

% =============================================================================
\section{Threat Model}
\label{sec:threat}

We study a binary email classifier $f_\theta : \mathcal{X} \rightarrow
\{0, 1\}$ that maps an email's concatenated subject and body
$x \in \mathcal{X}$ to a label $y$, where $y\!=\!1$ denotes phishing.
The defender deploys $f_\theta$ as part of an automated email filter and
seeks high accuracy across both ordinary phishing and adversarially
modified phishing.

\paragraph{Adversary capabilities.} We assume a \emph{black-box} text-level
adversary $\mathcal{A}$ who:
\begin{itemize}
  \item knows that the defender uses keyword- and URL-based features
        commonly cited in the literature, but does not have query access
        to the trained classifier or to training gradients;
  \item can rewrite the textual content of phishing emails with
        \emph{label-preserving} edits --- the message must still represent
        a credible phishing attempt to a human reader;
  \item cannot directly modify benign emails (this is consistent with the
        real attack surface: an attacker controls only the messages
        \emph{they themselves send}).
\end{itemize}
The adversary's goal is evasion: cause $f_\theta$ to predict $y\!=\!0$
(benign) on a phishing input.

\paragraph{Edit budget.} We constrain $\mathcal{A}$ to five categories of
label-preserving textual edits (Table~\ref{tab:ops}), drawn from
attacker behaviors documented in
\cite{kim2025llms,phishmaster2025,robustllmphish2025} and from the standard
adversarial-NLP toolkit~\cite{textattack2020}. Every edit either removes
a feature ($\mathit{mask\_urls}$, $\mathit{remove\_urgency}$,
$\mathit{rewrite\_subject}$) or substitutes a feature with a near-synonym
($\mathit{synonym\_swap}$, $\mathit{char\_perturb}$). No edit changes the
underlying semantic intent of the message.

\paragraph{Defender capabilities.} The defender knows the attacker's edit
budget at training time but does not see the attacker's actual perturbed
test inputs at training time. The defender's ``defense'' is to augment
the training data with one perturbed copy of each phishing email
(\S\ref{sec:augmentation}), obtained by sampling from the same edit
distribution. This is a non-adaptive defense: the defender does not
re-train against newly observed attacks.

\paragraph{Success criteria.} A defense succeeds if, relative to the
undefended baseline, it (i) does not significantly reduce accuracy on the
standard test set, and (ii) significantly improves accuracy on the
smart-attacker test set. ``Significantly'' is operationalized below as
``effect size larger than the std across three seeds.''

% =============================================================================
\section{Related Work}
\label{sec:related}

\subsection{Transformer-based phishing detection}
Atawneh and Aljehani~\cite{atawneh2023phishing} compare CNN, RNN, LSTM, and
BERT for email phishing classification, reporting up to 99.6\% accuracy with
BERT. More recent surveys and replications confirm that
BERT/RoBERTa/DistilBERT are competitive on multiple corpora, with DistilBERT
trading marginal accuracy for considerable inference-cost
savings~\cite{phishingbertroberta2025,indepthphishingdl2025,nadeepphishing2025}.
Alhogail and Alsabih~\cite{alhogail2021applying} additionally show that strong
detection is possible from email body text alone (without URLs or
attachments), motivating evaluation under URL removal --- but they do not
study other surface-cue removals such as urgency wording.

\subsection{LLM-generated and adversarial phishing}
Kim~et~al.~\cite{kim2025llms} document that LLM-crafted spear-phishing emails
achieve $\sim$47\% click rates, evading both human and software detection.
Phish-Master~\cite{phishmaster2025} explicitly uses chain-of-thought prompting
to generate phishing that bypasses enterprise filters. Robust ML-based
detection of LLM-generated and adversarial phishing using advanced
preprocessing has recently been shown to recover much of the lost
accuracy~\cite{robustllmphish2025}. Adversarial robustness of phishing
classifiers has also been benchmarked
directly~\cite{adversarialphishing2023,explainabletransformer2025}, with FGM
adversarial training raising BERT F1 from 86.2 to
91.7~\cite{explainabletransformer2025}. These results frame our work: the
attack surface is real, and lightweight defenses do help.

\subsection{Data augmentation for NLP robustness}
Junior et al.~\cite{junior2022dataaug} demonstrate the
augment--retrain--reevaluate pattern for fake-news detection. Recent surveys
emphasize that adversarial-style augmentation improves not only accuracy but
also generalization to out-of-distribution
inputs~\cite{textaugsurvey2025}. We use the
\textsc{TextAttack}~\cite{textattack2020} family of perturbation primitives
(synonym swap, character perturbations) as a starting point for our
smart-attacker pipeline. Recent attack methods such as PGD-based BERT
attacks~\cite{pgdbertattack2024} reinforce that BERT remains sensitive to
small, semantically preserving perturbations --- which is exactly what a
smart attacker would apply.

\subsection{Datasets and benchmarks}
We adopt the curated multi-source phishing dataset of
Champa~et~al.~\cite{champa2024curated}, which standardizes 11 publicly
available phishing/benign email corpora into a unified format and provides
feature-based ML baselines. Its multi-source structure lets us study
cross-source generalization in addition to robustness, mitigating the risk
that a model exploits source-specific artifacts.

% =============================================================================
\section{Method}
\label{sec:method}

\subsection{Task formulation}
Given an email represented by its concatenated subject and body
$x=\langle s, b \rangle$, the model predicts a binary label
$y\in\{0,1\}$ indicating benign~(0) or phishing~(1). We minimize the
cross-entropy loss over a fine-tuning corpus drawn from
Champa~et~al.~\cite{champa2024curated}, with stratified 80/10/10
train/validation/test splits.

\subsection{Backbone classifier}
We fine-tune \textbf{bert-base-uncased}~\cite{devlin2019bert} via the
HuggingFace \texttt{transformers} library~\cite{wolf2020transformers}.
The classifier head is a single linear layer over the \texttt{[CLS]}
representation, trained with cross-entropy loss. Hyperparameters are
fixed across all six runs (3 baseline seeds + 3 augmented seeds) and
listed in Table~\ref{tab:hp}.

\begin{table}[t]
\caption{Training hyperparameters. Identical across all six runs.}
\label{tab:hp}
\centering
\begin{tabular}{ll}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Backbone                & \texttt{bert-base-uncased} (110M params) \\
Max sequence length     & 256 tokens \\
Batch size (train)      & 64 \\
Batch size (eval)       & 32 \\
Epochs                  & 3 \\
Optimizer               & AdamW, $\beta\!=\!(0.9, 0.999)$ \\
Learning rate           & $2\!\times\!10^{-5}$ \\
Weight decay            & 0.01 \\
LR schedule             & Linear, 10\% warmup \\
Gradient clipping       & 1.0 ($\ell_2$ norm) \\
Mixed precision         & bfloat16 (A100) \\
Seeds                   & $\{42, 123, 2024\}$ \\
Augmentation factor $k$ & 1 (phishing rows only) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Smart-attacker augmentation}
\label{sec:augmentation}
We define five label-preserving edits that simulate a smart attacker
adapting a phishing email. Each is implemented in
\texttt{src/augmentation.py} and is applied only to phishing-class rows
(applying them to benign rows would only add noise to a class that does
not exhibit these cues). Table~\ref{tab:ops} summarizes the operations.

\begin{table}[t]
\caption{Smart-attacker augmentation operations.}
\label{tab:ops}
\centering
\begin{tabular}{p{0.22\columnwidth} p{0.66\columnwidth}}
\toprule
\textbf{Operation} & \textbf{Description} \\
\midrule
\texttt{mask\_urls} & Replace URLs with neutral placeholders (\texttt{[LINK]}), or with bare domain forms that look innocuous. \\
\texttt{remove\_} \texttt{urgency} & Remove or soften urgency keywords (\textit{URGENT}, \textit{ASAP}, \textit{immediately}, \textit{verify now}, \textit{suspended}, etc.) using a hand-curated lexicon and replacement table. \\
\texttt{rewrite\_} \texttt{subject} & Strip suspicious subject prefixes (\textit{URGENT:}, \textit{ALERT:}, \textit{ACTION REQUIRED:}) and remove emphatic punctuation. \\
\texttt{synonym\_} \texttt{swap} & Replace a small fraction of content words with WordNet synonyms ($p\!=\!0.1$, max 6 swaps per email). \\
\texttt{char\_} \texttt{perturb} & Light character-level noise (Cyrillic homoglyph substitution, adjacent-letter swap) at $p\!=\!0.02$. \\
\bottomrule
\end{tabular}
\end{table}

We use the augmentation pipeline in two distinct ways, summarized in
Algorithms~\ref{alg:smart} and~\ref{alg:augtrain}:

\paragraph{Smart-attacker test split} For each phishing email in the
held-out test set we apply the entire pipeline once. Benign emails are left
unchanged. The resulting split has the \emph{same} labels as the standard
test set; the only thing that changes is that phishing emails no longer
exhibit obvious surface cues. A robust detector should perform comparably on
both splits.

\begin{algorithm}[t]
\caption{Smart-attacker test split construction}
\label{alg:smart}
\begin{algorithmic}[1]
\Require Test set $D_{\text{test}}$, edit list $\mathcal{O}$
\Ensure  Smart-attacker test set $D_{\text{smart}}$
\State $D_{\text{smart}} \gets \emptyset$
\ForAll{$(x, y) \in D_{\text{test}}$}
  \If{$y = 1$}
    \Comment{phishing email — apply attacker pipeline}
    \State $x' \gets x$
    \ForAll{op $\in \mathcal{O}$}
      \State $x' \gets \textsc{op}(x')$
    \EndFor
    \State $D_{\text{smart}} \gets D_{\text{smart}} \cup \{(x', 1)\}$
  \Else
    \Comment{benign email — leave unchanged}
    \State $D_{\text{smart}} \gets D_{\text{smart}} \cup \{(x, 0)\}$
  \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

\paragraph{Train-time augmentation} For each phishing email in the training
set we generate $k\!=\!1$ additional augmented copy by passing it through the
pipeline. The augmented copies retain the phishing label and are concatenated
with the original training data. We then retrain the classifier from scratch.

\begin{algorithm}[t]
\caption{Train-time smart-attacker augmentation ($k\!=\!1$)}
\label{alg:augtrain}
\begin{algorithmic}[1]
\Require Training set $D_{\text{train}}$, edit list $\mathcal{O}$
\Ensure  Augmented training set $D_{\text{train}}^{\,+}$
\State $D_{\text{train}}^{\,+} \gets D_{\text{train}}$
\ForAll{$(x, y) \in D_{\text{train}}$ \textbf{with} $y = 1$}
  \State $x' \gets x$
  \ForAll{op $\in \mathcal{O}$}
    \State $x' \gets \textsc{op}(x')$ \Comment{stochastic op}
  \EndFor
  \State $D_{\text{train}}^{\,+} \gets D_{\text{train}}^{\,+} \cup \{(x', 1)\}$
\EndFor
\State \textbf{shuffle}~$D_{\text{train}}^{\,+}$
\end{algorithmic}
\end{algorithm}

\subsection{Evaluation protocol}
We compare three model stages: (S1)~baseline trained on the original data
and evaluated on the standard test set, (S2)~the same baseline evaluated on
the smart-attacker test set, and (S3)~the augmented model evaluated on both
splits. The expected pattern is a large drop from S1 to S2, partially closed
by S3. We additionally run a \emph{per-edit ablation}: each operation is
applied alone to the test phishing rows, isolating which surface cue the
baseline most depends on.

% =============================================================================
\section{Experimental Setup}
\label{sec:setup}

\subsection{Implementation}
The codebase is in PyTorch with HuggingFace
\texttt{transformers}~\cite{wolf2020transformers}. Training is driven by
\texttt{src/train.py}; evaluation, including the per-edit ablation, by
\texttt{src/evaluate.py}. The \texttt{notebooks/phishing\_pipeline.ipynb}
notebook reproduces all three stages on Google Colab. A pipeline-level
sanity test (\texttt{scripts/sanity\_check.py}) validates the augmentation
ops and the data-loader on synthetic data without requiring GPU access; a
tiny end-to-end smoke test (\texttt{scripts/tiny\_train\_test.py}) exercises
forward/backward/save/reload using a tiny CI-style BERT model.

\subsection{Dataset}
We use the figshare-hosted release of the Champa~et~al.\ curated phishing
email dataset~\cite{champa2024curated}.\footnote{\url{https://figshare.com/articles/dataset/Curated_Dataset_-_Phishing_Email/24899952}}
The data loader (\texttt{src/data\_loader.py}) is column-name agnostic so
that all CSV variants in the figshare bundle can be ingested with a single
script.

\subsection{Compute and reproducibility}
All training is performed on a Google Colab \textbf{A100} GPU with
\texttt{bfloat16} mixed precision (the codebase falls back to
\texttt{fp16}~+~\texttt{GradScaler} on T4/V100). Each configuration
(baseline, augmented) is trained with three seeds, $\{42, 123, 2024\}$,
controlling NumPy, PyTorch, and Python's \texttt{random} module. The same
train/val/test split (stratified, 80/10/10) is shared across seeds; only
model initialization, batch shuffling, and the random choices inside the
augmentation pipeline differ between seeds. We report mean~$\pm$~std
across the three seeds. After deduplication and column-normalization, the
combined corpus contains $N\!=\!131{,}171$ emails (phishing $67{,}230$;
benign $63{,}941$). Two source files
(\texttt{TREC\_05.csv}, \texttt{TREC\_06.csv}) failed to parse due to
malformed rows and were dropped, leaving five usable corpora (CEAS-08,
Enron, Ling, SpamAssassin, TREC-07). The augmentation pipeline used in
both train-time augmentation and the smart-attacker test split contains
five operations: \texttt{mask\_urls}, \texttt{remove\_urgency},
\texttt{rewrite\_subject}, \texttt{synonym\_swap}, and
\texttt{char\_perturb}.

% =============================================================================
\section{Results}
\label{sec:results}

Table~\ref{tab:headline} presents the headline four-way comparison: each of the
two models (baseline, augmented) evaluated on each of the two test splits
(standard, smart-attacker). Detailed metrics for the augmented model are in
Table~\ref{tab:augfull}, and the per-edit ablation under the augmented model
is in Table~\ref{tab:ablation}.

\begin{table}[t]
\caption{Headline results across three seeds (mean~$\pm$~std). The vanilla
baseline is essentially insensitive to the smart-attacker edits ($\Delta$
within seed noise). Augmentation produces a clean $+0.0020$ F1 gain on the
smart-attacker test at zero cost on the standard test.}
\label{tab:headline}
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{Standard F1} & \textbf{Smart F1} & $\Delta$ \\
\midrule
S1: BERT baseline             & 0.9958\,$\pm$\,0.0002 & 0.9957\,$\pm$\,0.0001 & $-0.0001$ \\
S2: BERT + aug.\ training      & 0.9958\,$\pm$\,0.0003 & 0.9977\,$\pm$\,0.0001 & $+0.0019$ \\
\midrule
$\Delta$ (S2$-$S1)            & $+0.0000$              & $+0.0020$              & --- \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[t]
\caption{Augmented model, full per-split metrics across three seeds
(mean~$\pm$~std). Recall on the phishing class is the primary mover: the
augmented model becomes more confident on phishing exactly when surface
cues are stripped.}
\label{tab:augfull}
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Metric} & \textbf{Standard} & \textbf{Smart-attacker} \\
\midrule
Accuracy        & 0.9957\,$\pm$\,0.0003 & 0.9976\,$\pm$\,0.0001 \\
F1 (phishing)   & 0.9958\,$\pm$\,0.0003 & 0.9977\,$\pm$\,0.0001 \\
Recall (phish.) & 0.9957\,$\pm$\,0.0006 & 0.9995\,$\pm$\,0.0002 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\caption{Per-edit ablation on the augmented model across three seeds
(mean~$\pm$~std). Each row applies one smart-attacker edit alone to phishing
emails; benign emails are unchanged. \texttt{char\_perturb} is the
single most impactful edit; \texttt{rewrite\_subject} is essentially a no-op
because the subject-prefix regex rarely fires on this corpus.}
\label{tab:ablation}
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Edit applied alone} & \textbf{F1 (phishing)} & \textbf{Recall (phishing)} \\
\midrule
\textit{None (standard test)}      & 0.9958\,$\pm$\,0.0003 & 0.9957\,$\pm$\,0.0006 \\
\texttt{mask\_urls}                & 0.9963\,$\pm$\,0.0001 & 0.9966\,$\pm$\,0.0004 \\
\texttt{remove\_urgency}           & 0.9958\,$\pm$\,0.0004 & 0.9957\,$\pm$\,0.0008 \\
\texttt{rewrite\_subject}          & 0.9958\,$\pm$\,0.0003 & 0.9956\,$\pm$\,0.0007 \\
\texttt{synonym\_swap}             & 0.9964\,$\pm$\,0.0002 & 0.9968\,$\pm$\,0.0002 \\
\texttt{char\_perturb}             & 0.9975\,$\pm$\,0.0001 & 0.9992\,$\pm$\,0.0001 \\
\textit{All combined (smart test)} & 0.9977\,$\pm$\,0.0001 & 0.9995\,$\pm$\,0.0002 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar, ymin=0.992, ymax=0.999,
  width=0.95\columnwidth, height=4.6cm,
  bar width=12pt,
  enlarge x limits=0.35,
  ylabel={F1 (phishing)},
  symbolic x coords={Standard, Smart-attacker},
  xtick=data,
  legend style={at={(0.5,-0.22)}, anchor=north, legend columns=2,
                font=\footnotesize, /tikz/every even column/.append style={column sep=10pt}},
  ymajorgrids, grid style={dashed,gray!30},
  every node near coord/.append style={font=\tiny, /pgf/number format/.cd, fixed, precision=4},
  error bars/y dir=both, error bars/y explicit,
]
\addplot+[fill=blue!25] coordinates {
  (Standard, 0.9958) +- (0, 0.0002)
  (Smart-attacker, 0.9957) +- (0, 0.0001)
};
\addplot+[fill=orange!40] coordinates {
  (Standard, 0.9958) +- (0, 0.0003)
  (Smart-attacker, 0.9977) +- (0, 0.0001)
};
\legend{Baseline, Augmented}
\end{axis}
\end{tikzpicture}
\caption{Phishing-class F1 by model and test split, mean $\pm$ std across
three seeds. The baseline is essentially flat across splits; augmentation
lifts the smart-attacker bar without moving the standard bar.}
\label{fig:headline}
\end{figure}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar, ymin=0.993, ymax=0.999,
  width=0.95\columnwidth, height=4.8cm,
  bar width=12pt,
  enlarge x limits=0.10,
  ylabel={F1 (phishing)},
  symbolic x coords={none, mask, urgency, subject, synonym, char, all},
  xtick=data,
  xticklabel style={font=\scriptsize, rotate=20, anchor=east},
  ymajorgrids, grid style={dashed,gray!30},
  error bars/y dir=both, error bars/y explicit,
]
\addplot+[fill=teal!35] coordinates {
  (none, 0.9958) +- (0, 0.0003)
  (mask, 0.9963) +- (0, 0.0001)
  (urgency, 0.9958) +- (0, 0.0004)
  (subject, 0.9958) +- (0, 0.0003)
  (synonym, 0.9964) +- (0, 0.0002)
  (char, 0.9975) +- (0, 0.0001)
  (all, 0.9977) +- (0, 0.0001)
};
\draw[dashed, gray] (axis cs:none, 0.9958) -- (axis cs:all, 0.9958);
\end{axis}
\end{tikzpicture}
\caption{Per-edit ablation on the augmented model (mean $\pm$ std, 3 seeds).
\texttt{char} = \texttt{char\_perturb} dominates; \texttt{urgency} =
\texttt{remove\_urgency} and \texttt{subject} = \texttt{rewrite\_subject}
are no-ops within noise (dashed line = standard-test F1).}
\label{fig:ablation}
\end{figure}

\paragraph{Headline finding.} Table~\ref{tab:headline} shows that, averaged
over three seeds, the vanilla BERT baseline (S1) is essentially insensitive
to the smart-attacker edits: F1 changes by $-0.0001$ between the standard
and smart-attacker tests, and $0.0001$ is also the std of either cell, so
the change is well within seed noise. The over-reliance hypothesis we set
out to test is therefore not supported on this corpus. After training with
smart-attacker augmentation (S2), the model gains $+0.0020$ F1 on the
smart-attacker test at a cost of $+0.0000$ F1 (i.e.~no measurable cost) on
the standard test. The std on the augmented smart-attacker cell is just
$0.0001$, so this $+0.0020$ gain is not attributable to seed variance.

\paragraph{What augmentation actually changes.} Augmentation does not
``recover'' performance the baseline lost --- the baseline never lost
anything. Instead, it raises the ceiling on the smart-attacker split. On
the augmented model (Table~\ref{tab:augfull}), recall on the phishing class
moves from $0.9957$ on the standard split to $0.9995$ on the smart-attacker
split: the model becomes \emph{more} confident on phishing exactly when its
easily-memorized features (URLs, urgency tokens) are removed. We read this
as evidence that the augmented model has learned to classify from semantic
context. False positives (the precision side) remain stable, since
augmentation only touches phishing rows and leaves benign emails unchanged.

\paragraph{Per-edit ablation.} Table~\ref{tab:ablation} shows the
contribution of each smart-attacker edit applied alone to the augmented
model. Three observations:
\begin{itemize}
  \item \texttt{char\_perturb} is the single largest contributor
        (F1 = 0.9975, recall = 0.9992), beating \texttt{mask\_urls} and
        \texttt{synonym\_swap}. The augmented model is particularly
        confident on phishing when light character noise is applied, likely
        because the augmentation pipeline trained it to be robust to
        homoglyph and adjacent-letter perturbations.
  \item \texttt{rewrite\_subject} and \texttt{remove\_urgency} are
        effectively no-ops, with means within $\pm 0.0001$ of the
        standard test. \texttt{rewrite\_subject} is mechanical: the
        subject-prefix regex (\texttt{URGENT:}, \texttt{ALERT:},
        \texttt{ACTION REQUIRED:}, \ldots) rarely fires on this corpus.
        \texttt{remove\_urgency} fires on every row but the augmented model
        is already invariant to those tokens.
  \item The combined-edit row exceeds every single edit, suggesting
        the gains are weakly additive: each edit removes a different
        surface dependency that the augmented model is independently
        robust to.
\end{itemize}

% =============================================================================
\section{Qualitative Analysis}
\label{sec:qualitative}

To make the results in \S\ref{sec:results} concrete, we walk through three
representative phishing emails from the test set, comparing the original
text to the smart-attacker variant produced by the full pipeline.\footnote{
Emails are lightly redacted/shortened for space. Full unredacted versions
are reproduced verbatim by \texttt{notebooks/phishing\_pipeline.ipynb}.}

\subsection{Example 1: account-verification phish}

\noindent\textbf{Original (label = phishing).}
\begin{lstlisting}
Subject: URGENT: Verification required
Your Amazon account is subject to deletion
unless you verify immediately at
https://amaz0n-verify.example.com/login.
Act now!
\end{lstlisting}
\noindent\textbf{Smart-attacker variant.}
\begin{lstlisting}
Subject: Verification required
Your Amazon report is open to deletion
unless you verify soon at [LINK].
please respond
\end{lstlisting}

\noindent The smart-attacker version retains the recognizable
\textit{phishing topic} (account verification, threatened deletion,
external link) but loses every surface marker the proposal predicted
would matter: the \texttt{URGENT:} subject prefix is gone, the URL is
masked, ``immediately'' has become ``soon,'' and ``Act now!'' has become
``please respond.'' Both the baseline and augmented models classify both
versions as phishing in our experiments, consistent with the headline
result.

\subsection{Example 2: bulk pharmacy spam}

This is a representative member of the dominant class in the corpus
(legacy bulk spam from CEAS-08 and Enron). The message is on-topic for
phishing in this dataset's labeling convention but does not look anything
like a modern targeted phishing attempt.

\noindent\textbf{Original (label = phishing).}
\begin{lstlisting}
Subject: Psychologist endorsed products for men
Smell better with the girls with a stronger
pheromone product. http://shop-rx.example/login
-- Use Opera's revolutionary e-mail client.
\end{lstlisting}
\noindent\textbf{Smart-attacker variant.}
\begin{lstlisting}
Subject: Psychologist endorsed products for men
Smell better with the girls with a stronger
pheromone product. [LINK]
-- Utilization Opera's revolutionary e-mail client.
\end{lstlisting}

\noindent The smart-attacker variant differs only in the masked URL and a
synonym swap (\textit{Use}~$\rightarrow$~\textit{Utilization}). Notice
that the surrounding text remains overwhelmingly indicative of the
phishing label even after edits --- it is precisely this corpus property
that explains the null baseline-drop result of \S\ref{sec:results}. A
classifier with any reasonable representation of bulk-spam topics would
classify this message identically before and after augmentation.

\subsection{Example 3: news-headline lure}

\noindent\textbf{Original (label = phishing).}
\begin{lstlisting}
Subject: CNN.com Daily Top 10
TOP 10 VIDEOS
1. PARIS HILTON TAKES ON MCCAIN
   http://cnn.example/p1.html
2. BIKINI BARISTA STAND CLOSED
3. MAGGOTS: THE NEW ANTIBIOTIC?
   http://cnn.example/p3.html
...
\end{lstlisting}
\noindent\textbf{Smart-attacker variant.}
\begin{lstlisting}
Subject: [Colligate] Casual Top 10
TOP 10 VIDEOS
1. PARIS HILTON TAKES ON MCCAIN [LINK]
2. BIKINI BARISTA STAND CLOSED
3. MAGGOTS: THE NEW ANTIBIOTIC? [LINK]
...
\end{lstlisting}

\noindent Here the synonym-swap operator produces a notably non-fluent
title (\textit{CNN.com Daily}~$\rightarrow$~\textit{[Colligate] Casual}).
This is one of the failure modes of WordNet-based synonym swap on
named-entity tokens. Despite the disfluency, the rest of the message
remains classifiable; we have not observed cases where synonym swap alone
flips the model's prediction.

\subsection{Take-away}

The qualitative pattern matches the quantitative results: removing
surface cues from this corpus's phishing emails leaves enough signal in
the surrounding content for both the baseline and augmented model to
classify correctly. We expect this margin to shrink on a corpus where the
original phishing emails read more like ordinary business correspondence
(e.g.\ modern LLM-generated phishing).

% =============================================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Why the baseline does not drop.} The proposal anticipated a
substantial accuracy drop from S1's standard test to S1's smart-attacker
test, on the basis that BERT would over-rely on surface tokens such as
\textit{URGENT}, \textit{verify}, and explicit URL substrings. With a
single seed we observed a $0.10$\,pp F1 drop and provisionally accepted a
weak version of the hypothesis. With three seeds the drop disappears
entirely (mean $-0.0001$ F1, std $0.0001$). The hypothesis is therefore
\emph{not} supported on this corpus. We attribute this null result to two
properties of the corpus and the model.

First, the Champa~et~al.\ release combines five public corpora
(CEAS-08, Enron, Ling, SpamAssassin, TREC-07) whose phishing labels include
a large amount of \emph{classical bulk spam} --- pharmacy advertisements,
lottery scams, adult-content solicitations --- whose label is supported by
many redundant features (topic, register, lexical density of advertising
language) in addition to the URLs and urgency keywords our pipeline edits.
Removing surface cues from these messages does not flip the label because
the surrounding text continues to read unambiguously as spam.

Second, BERT does not learn a sparse rule like ``\textit{contains
\textbf{URGENT}}~$\Rightarrow$~\textit{phishing}''; it learns a dense
contextual representation. A 3-epoch fine-tune on $\sim$100k emails
appears sufficient for the model to distribute classification weight
across many lexical and stylistic signals, leaving it relatively
insensitive to the loss of any single one.

A more decisive test of the over-reliance hypothesis would use
LLM-rewritten phishing of the kind reported by Phish-Master
\cite{phishmaster2025} or Kim~et~al.~\cite{kim2025llms}, where the entire
message --- not just URLs and urgency --- has been rewritten to read as
benign business correspondence.

\paragraph{What augmentation actually buys.} The headline gain from
augmentation is $+0.0020$ F1 on the smart-attacker test at $+0.0000$ F1
on the standard test --- a Pareto improvement, not a tradeoff. Standard
deviations across three seeds are $\le 0.0003$ in every cell of
Table~\ref{tab:headline}, so the effect is comfortably outside seed noise.
The result is internally consistent with Table~\ref{tab:ablation}, where
\texttt{char\_perturb} produces the largest single-edit gain even though
it is the perturbation type the proposal least expected to matter: the
augmentation procedure has trained the classifier to be invariant to
homoglyph and adjacent-letter noise that the baseline never explicitly
encountered.

A useful framing is that augmentation is not ``defensive'' in the
classical adversarial-training sense (the baseline was not under attack to
begin with); it is a regularizer that shifts the model's representation
toward more invariant features, with the side effect of further raising
F1 on perturbed inputs. The direction matches recent results in the
text-augmentation literature
\cite{textaugsurvey2025,explainabletransformer2025,robustllmphish2025},
though our effect size is small because the Champa corpus's phishing
signal is already easy.

\paragraph{Limitations.}
\begin{itemize}
  \item \emph{Vintage of corpus.} The Champa~et~al.\ collection draws on
        public corpora that are largely pre-2015. Modern, LLM-generated
        phishing exhibits qualitatively different surface statistics
        \cite{kim2025llms,phishmaster2025} and would be the more important
        target for a robustness study; our pipeline can be re-applied
        unchanged once such a corpus is publicly available.
  \item \emph{Finite edit vocabulary.} Our smart-attacker simulation is
        rule-based and therefore has a small, fixed set of perturbation
        types. A stronger evaluation would integrate gradient-based or
        beam-search adversarial attacks from frameworks such as
        TextAttack~\cite{textattack2020} or PGD-BERT-Attack
        \cite{pgdbertattack2024}, or LLM-rewritten phishing in the spirit
        of Phish-Master~\cite{phishmaster2025}.
  \item \emph{Phishing-only augmentation.} Augmentation is restricted to
        the phishing class. This was chosen so the smart-attacker semantics
        are well-defined (a smart attacker rewrites their phishing email,
        not the victim's reply), but it does shift the class prior of the
        training mix slightly. We mitigate this with a small augmentation
        factor ($k\!=\!1$); the constant standard-test F1 across S1 and
        S2 (Table~\ref{tab:headline}) suggests the shift was not
        consequential at this scale.
  \item \emph{Subject-rewrite regex.} Our \texttt{rewrite\_subject} regex
        targets prefixes (\texttt{URGENT:}, etc.) that are uncommon in the
        Champa subjects, hence the no-op row in
        Table~\ref{tab:ablation}. A more aggressive subject-line
        transformation (e.g.\ LLM paraphrase) would likely change this row.
  \item \emph{Single train/val/test split.} Across the three seeds we hold
        the same data split fixed and vary only model initialization,
        batch order, and augmentation random sampling. This isolates
        \emph{model} variance but not \emph{data} variance; a fuller study
        would use multiple resampled splits as well.
\end{itemize}

% =============================================================================
\section{Future Work}
\label{sec:future}

We see four concrete directions in which the present study should be
extended.

\paragraph{1. Modern LLM-generated phishing.} The most important
follow-up is to re-run the same pipeline on a corpus of \emph{modern},
LLM-generated phishing such as those described by Kim~et~al.\ and the
Phish-Master framework~\cite{kim2025llms,phishmaster2025}. We expect that
the redundancy property which makes our baseline robust on Champa~et~al.\
will not hold on LLM-rewritten phishing, in which case the baseline drop
should grow substantially and augmentation's role would shift from
``free regularizer'' to ``necessary defense.'' The codebase already
exposes this as a single configuration change; the bottleneck is data
availability.

\paragraph{2. Stronger adversarial attacks.} Replacing our rule-based
edit pipeline with a gradient-based or search-based attack
(BAE~\cite{textattack2020}, PGD-BERT-Attack~\cite{pgdbertattack2024})
would let the attacker discover model-specific failure modes that our
fixed lexicon cannot. We anticipate larger baseline drops here even on
Champa~et~al., because the attack would be optimized against the
classifier rather than against a generic surface-feature heuristic.

\paragraph{3. Adaptive defenses.} Our augmentation is non-adaptive: the
defender samples once from a fixed edit distribution. An adaptive
counterpart would alternate attack and defense rounds, in the spirit of
classical adversarial training~\cite{explainabletransformer2025}, and we
expect this to widen the augmented model's lead on under-attack F1
without reopening clean-data degradation.

\paragraph{4. Cross-source generalization.} The Champa~et~al.\ release
combines multiple public corpora; our experiments collapse them. A
\emph{leave-one-corpus-out} evaluation, where the model is trained on
four corpora and tested on the held-out fifth, would reveal whether
the apparent robustness here is a property of BERT or simply of in-source
generalization. Our data loader already tags each row with its source
corpus, so this is a reformulation of \texttt{src/data\_loader.py} rather
than new code.

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

We presented an experimental pipeline for studying the robustness of
Transformer-based phishing email classifiers under simulated ``smart
attacker'' edits. The pipeline includes (i) a label-preserving augmentation
module covering URL masking, urgency removal, subject rewriting, synonym
swap, and character perturbation; (ii) a held-out smart-attacker evaluation
split; (iii) a per-edit ablation that localizes which surface cue the model
relies on; (iv) an augmented-training counterpart that quantifies how much
extra robustness can be obtained; and (v) a multi-seed aggregator that
reports mean~$\pm$~std across seeds.

Across three seeds on the Champa~et~al.\ multi-source corpus, vanilla
\textsc{BERT-base} is essentially \emph{perfectly} robust to our edits:
phishing F1 changes by $-0.0001$ under the smart-attacker test, within
seed noise. Targeted augmentation does not need to defend a broken
baseline, but still produces a clean Pareto improvement: $+0.0020$ F1 on
the smart-attacker test at $+0.0000$ F1 on the standard test, with
\texttt{char\_perturb} as the most impactful single edit in the per-edit
ablation. The over-reliance hypothesis we set out to test is therefore
\emph{not} supported on this corpus; we attribute the null result to
redundancy in the corpus's phishing signal and to BERT's distributed
representation, and expect the same pipeline to produce considerably
larger baseline drops on modern LLM-generated phishing. We have packaged
the codebase to make that re-evaluation a single command-line flag away.

\section*{Code and data}
The complete codebase, Colab notebook, and LaTeX source for this manuscript
are available in the project repository (see \texttt{README.md}).

\section*{Acknowledgments}
We thank the CSE 587 instructors and TAs at Penn State for guidance, and
Champa et al.\ for releasing the curated phishing email dataset that made
this study possible.

% =============================================================================
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
